### 2024-07-19 rebuttals/rebuttals.tex

\noindent In my opinion, the main variables of EnergyScope Pathway are presented {\color{blue} in Section 1.1 of the core of the manuscript}. Further details are also given {\color{blue}in Appendix B.2}. Finally, these variables are even exhaustively detailed in \cite{limpens2019energyscope} and \cite{limpens2024pathway} that are cited several times in the thesis manuscript.  Regarding the uncertainties, the way the ranges are defined and the way they affect the value of some parameters in the model are explained {\color{blue}in Section 1.2 of the core of the manuscript}. Consequently, no modification has been brought to the manuscript regarding this topic.

\subsection{Assumptions - Fine tuning}
\label{fine_tuning}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}}, {\color{purple} \textbf{Sylvain}} \& {\color{teal} \textbf{Stefan}} - Many assumptions, e.g., alpha = 85\% (p. 40) or reward = -300 (Chapter 4). How sensitive are the results to these assumptions? It looks like all these parameters have been tuned to obtain the desired results. How robust are they? Can the methodology be generalized to other models? Can we keep the same values for other applications? You need a model that runs pretty fast, right?

\noindent {\color{teal} \textbf{Stefan}} - Justification of the RL actions selected? why no other actions like the cost of some technologies?

\noindent {\color{purple} \textbf{Sylvain}} - p. 86 - A lot of heuristic and trial and error. Can the methodology be generalized to other models?

\noindent {\color{purple} \textbf{Sylvain}} - Make sure to discuss the hypotheses because one might not agree with all of them. So, it could provide more context but also some limitations and indications of the alternatives (and the spectrum of what is possible).
\end{mdframed}

@@

\noindent I have not carried out analysis to assess the sensitivity of the results to these assumptions. They result either from trial and error or observations made \textit{a posteriori}. However, to discuss these assumptions and the methodological choices, I have added a specific section to each methodological parts of Chapter 1 to provide future users and developers of the methods proposed in this work with guidelines. These aim at helping other researchers in either using the same methodologies and assumptions on their own case study or keep on developing and improving these methodologies.

\begin{mdframed}[style=manuscript]
\textbf{1.1.3 Discussion and guidelines for future researchers} [EnergyScope Pathway]

Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix B.1).

@@

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
\textbf{4.2.4 Discussion and guidelines for future researchers}

As introduced in Chapter 1, most of the work in applying RL is the definition of the interactions between the agent and its environment (\ie actions, reward and states) that are very dependent on the case study and the research questions to answer. The elements presented in this work result from several trial and errors to end up with meaningful results according to our research questions.

Besides mimicking potential actual policies, the actions chosen in this work have a direct translation into constraints and we can therefore assess their effectiveness through the fact they are binding or not. In this work, we have investigated other actions like incentivising solar PV panels and wind turbines by ``artificially'' reducing their CAPEX. The result was not conclusive as these technologies need to take part to the Belgian energy transition and because it was harder to assess the impact of this action.

The reward function was designed to first aiming at respecting the \ce{CO2} budget then minimising the total transition cost. The -300 penalty given in case of an infeasible optimisation problem was arbitrarily set. \textit{A posteriori}, it seems to be a well-defined penalty given its significant relative difference with the values taken by the reward otherwise, between -120 and 44 (see Figure 4.4). Besides this penalty and given the observed results, we recommend to start with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs. This requires to define the \ce{CO2} budget according to a certain sharing principle (see Section 2.5) and to compute reference total transition cost. However, other researches focusing on reaching carbon neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.

@@

\subsection{Graphs and results}
\label{general_graphs_results}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Many axes labels are missing, and the captions are often insufficient to read the figures. 
\end{mdframed}

\noindent I agree and have gone through all the figures of the manuscript to clarify them by adding axes labels or extending the caption where it was needed. Figures where y-axis (and its label) is missing are distribution (or density) plots (\ie Figures 3.6, 4.4, 4.5, 4.6, 4.8 and E.1).  Consequently, similarly to an histogram, the feature of interest and its span of values are represented on the x-axis. The y-axis has no real interest in terms of information. What matters is the shape of the curve to be able to point out for which values/ranges of values, the feature of interest occurred more or less often.  For this reason, I have not added y-axis for the aforementioned figures but rather improved their respective caption, emphasising, among other things, that these figures represent distribution of occurrences. For instance, the caption of {\color{blue}Figure 4.4} is now more explicit:

\begin{mdframed}[style=manuscript] % Comment from the reviewer
Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right). Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050. Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest. The majority of successful transitions have cumulative emissions much lower than the \ce{CO2} budget and are cheaper than the REF case. 
\end{mdframed}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{teal} \textbf{Stefan}} - Some big topics are missing like storage. Why is storage not more discussed? 
\end{mdframed}

\noindent As discussed during the private defense, storage is not more discussed in the manuscript for two reasons. First, the storage technologies, for the case of Belgium assuming a copper-plate approach, consist in two main technologies: methane storage mostly (about 15\,TWh) and, to a smaller extent, seasonal thermal storage (about 3\,TWh), with limited variations through the transition (see Figure \ref{fig:STORAGE_2}). Analyses under uncertainties have not shown major variations neither. Second, storage assets are intrinsically linked to the conversion technologies and, consequently, can be seen as by-products of the optimisation results. For these two reasons, I have rather focused my analyses on the use of resources and the installation and operation of conversion technologies. Given the discussion during the private defense and these further explanations, I have not brought modification to the manuscript in this regard.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\textwidth]{STORAGE_2.pdf}
\caption{Storage installed capacities for the REF case}
\label{fig:STORAGE_2}
\end{figure}

As discussed during the private defense, the first key-message of this work is the need for this (mix of) ``unicorn'' solution(s) given the ambitious objectives set to mitigate climate change.  This need for this radical change and the fact that solutions rely a lot on assumed to be carbon neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
Given the ambitious \ce{CO2} budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes. Assumed to be carbon neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition. This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
\end{mdframed}

@@

\noindent
It is an interesting topic for further investigations, especially how they address the subject in other parts of the world like Australia/Asia. This could be linked with the political perception of energy system optimisation models and the subsequent analyses. However, no modification has been brought to the manuscript regarding this topic.

\subsection{Writing style}
\label{writing_style}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - While a more ``friendly'' writing style can be pleasant, there are many colloquial expressions that not do not fit well in the thesis, \eg ``carrot and stick'', ``apples with apples'', etc. I recommend reducing them to the bare minimum, or completely getting rid of them. 
\end{mdframed}

\noindent The listed colloquial expressions have been transformed and, overall, I have reviewed the manuscript to reduce them. For instance, {\color{blue} in the contributions of Chapter 2}, I have changed ``lion's share'' into ``bigger share''.

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - Often, multi-stage models are affected by the end-of-time-horizon effects. Did you observe those and how did you deal with them? See, e.g. Figure 3.3 p. 66.
\end{mdframed}

 \noindent To cope with these ``end-of-time-horizon effects'', we have implemented the salvage value, $\textbf{C\textsubscript{inv,return}}$. This represents the residual investment that would remain after 2050. As detailed in Section 1.1.1: ``the salvage value avoids penalising the capital intensive technologies towards the end of the transition'' \cite{poncelet2016myopic}. The sensitivity analysis on the formulation of the salvage value has been added in Appendix B.2.2 (see further comments on the salvage value).

As discussed during the private defense, instead of implementing a salvage value another alternative would be to implement artificial representative years after 2050 that are then excluded from the total transition cost and emissions in while post-processing the results of the optimisation. Even though the choice of the formulation of the salvage value accounts for assumptions (discussed in Appendix B.2.2), we opted for this rather than ``artificial future years'' for several reasons. First of all, it allows mitigating the computational burden of the pathway problem. Indeed, the more there are representative years, the bigger are the matrices and the longer it takes to solve the problem. Second, the salvage value-approach is also used in other works \cite{poncelet2016myopic,prina2019transition}. Finally, implementing ``artificial  future years'' would require to either collect more data, which is quite time-consuming, or make other kind of assumptions on the evolution of costs, GWP and other features beyond 2050.

Since the implementation of the salvage value and the formulation choices are the work developed by \citet{limpens2024pathway} (partly integrated in Appendix B) and given the above explanations, I have not brought further modification to the manuscript in this regard.

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - p.17-19: is the myopic version modeled as a single LP or as a sequence of LPs? This should be better clarified. The diagram at p. 19 focuses on the code, it would be better replaced by a code showcasing the logic of the approach.
\end{mdframed}

\noindent Even though this sequence of LPs was already mentioned {\color{blue}in the caption of Figure 1.3}, I have complemented the text {\color{blue} in the paragraph ``Myopic pathway implementation'' of Section 1.1.2},

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
[...] The myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window, until eventually reaching 2050.
\end{mdframed}

@@

\noindent reminded it in the caption of Figure 1.2

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
The myopic approach (in pink) uses several instances of the pathway model, illustrated in Figure 1.1, as a sequence of different and linked LP problems. 
\end{mdframed}

\noindent and extended the caption of {\color{blue} Figure 1.4}. However, besides highlighting the key step 3.3 where there is an update of parameters value, I have decided to keep the figure as is since, on the one hand, the extension of the caption provides now a better understanding and, on the other hand, the code itself is exactly what is schematically represented on the diagram.

\begin{figure}[htbp!]
\centering
\includegraphics[width=10cm]{MY_process_code.pdf}
\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway. Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window. If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
\label{fig:MY_process_code}
\end{figure}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{teal} \textbf{Stefan}} - With the myopic approach, are you predicting what will come or drawing a path that is the best? Is that a desirable feature to mimic the way we proceed now. Why artificially limit the model as we know we have to go to zero in 2050?
\end{mdframed}

\noindent As written in the manuscript and discussed during the private defense, the purpose of the myopic approach is to challenge the assumption of the complete knowledge about the uncertain future and mimic the short-sightedness of the decision-making process. This sequential optimisation allows showcasing robust policies to support myopic transitions (see Chapter 4) and the robustness of roadmaps (see Chapter 5). The myopic approach is also the methodological basement to support the implementation of the RL and PCA-based methods.

Given these explanations and what is already in the manuscript, I have not brought modification in this regard.

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{teal} \textbf{Stefan}} - p.18 : Surprising that myopic leads to a sooner investment to the transition? Is it an effect of the salvage value? was there a sensitivity analysis on this salvage value? Perhaps include more details on this. The salvage value plays a large role in the decision of the model. Give more details to give more confidence in the choice.
\end{mdframed}

@@

\noindent Indeed, the discount rate is usually between 7.5\% and 12\% in other studies. The main justification for this choice was to be in line with G. Limpens' work \cite{limpens2021generating}. However, assessing the impact of $i_{\text{rate}}$ on $\tau_{phase}$, we see that a lower value of $i_{\text{rate}}$ will favour less late expenses compared to a high value. In other words, considering $i_{\text{rate}}=$1.5\% rather than 10\% will less postpone investments (see Figure \ref{fig:tau_phase}). I have added some explanations in this regard {\color{blue}in Section 1.1.1}:

\begin{figure}[htbp!]
\centering
\includegraphics[width=10cm]{tau_phase.png}
\caption{$\tau_{phase}$ versus the timing of the expenses (columns) and the discount rate, $i_rate$ (rows).}
\label{fig:tau_phase}
\end{figure}

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
In EnergyScope Pathway, $\tau\textsubscript{\emph{phase}}$ aims at depreciating the value of expenses done in the future. The discount rate accounted for in this annualisation factor, $i_{\text{rate}}$, is considered as identical for all the technologies of the system. In line with \citet{limpens2021generating}, we set the nominal value to $i_{\text{rate}}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}. Having a low value of discount rate favours less late expenses compared to a higher value of discount rate.
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - Can you explain the logic of the annualization factor in the pathway model?

\noindent {\color{purple} \textbf{Sylvain}} - p.13: Why no annualizing over the lifetime of the unit? Would that lead to different results? 
\end{mdframed}

@@

\noindent See previous comments.

\subsubsection{Power grid and transport infrastructure}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{purple} \textbf{Sylvain}} - How do you model the power grid interconnections with the neighbouring countries as well as the transport network of the different energy carriers within the country? p.63 - The model decides to use ammonia instead of SNG $\rightarrow$ is the transport infrastructure accounted for? p.73 - How is the transport infrastructure accounted for?
\end{mdframed}

\noindent Regarding the power grid interconnections with the neighbouring countries, similarly to the import of other energy carriers, there is no representation of those physical routes. The whole-energy system of the country is represented as a single-node and the imported energy commodities are available ``at the front door'' of the system. It goes the same way for the infrastructures to transport energy carriers within the system. As it is represented assuming a copper-plate approach, all the demands, resources and conversion technologies are concentrated in one single node. This is now explicitly written {\color{blue}in a new paragraph of Section 2.2}:

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - Can you explain the choice of N=5 for the uncertainty ranges?
\end{mdframed}

\noindent As detailed by \citet{Moret2017PhDThesis}: ``the ranges resulting from the uncertainty characterisation, are defined for a specific point in time in the planning horizon, usually its last year. Therefore, they need to be scaled accordingly when used in the optimisation model. Without this scaling, it would mean assuming that the uncertainty for year $N$ stands for all the years in the planning horizon''. In his thesis, Moret considered $N=20$ as the duration of the planning horizon between 2015, the year of Moret's work, and the snapshot optimisation of the Swiss case study in 2035. In my case, the planning horizon is 5-year long as the duration of a phase between two representative years of the transition. Referring to \citet{Moret2017PhDThesis} for the interested reader, I have explained a bit more the reason for this scaling factor {\color{blue}in the second paragraph of Section 2.4}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined ranges of uncertainties for the model parameters.  For the operation-type parameters (types II and III), a scaling factor needs to be considered to avoid assuming that the uncertainty for the last year of the planning horizon ($N$) is representative for all years in the planning horizon \citet{Moret2017PhDThesis}. In EnergyScope Pathway, as the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III. 
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - p. 20 - ``inspired by Guevara et al.'' $\rightarrow$ how was their approach used in your work?
\end{mdframed}

@@

\section{Chapter 2 - Case study}
\label{case_study}

\subsection{General and introduction}
\label{methodo_general}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - I am a bit underwhelmed by the many nitty-gritty data in this Chapter. While it is surely important to document everything, couldn’t at least part of this be moved to the Appendix?
\end{mdframed}

\noindent Most of the data about the Belgian case study have been collected by \citet{limpens2021generating}. The objective of this Chapter was to focus on my contributions: non-energy demand (NED), electrofuels, SMR, uncertain ranges for the pathway model and \ce{CO2} budget. However, to put these contributions in perspective with similar features in the model, I have decided to keep general information about the end-use demands, the resources and the conversion technologies. Consequently, I have kept the same level of information in Chapter 2 rather than in appendix.

### 2024-07-19 appendices/app_RL_TD_MO.tex
Rebuttals: Keep going...
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

The \acrfull{RL}-based exploration of myopic transition pathways can require thousands of runs to converge. To reduce the computational burden and maximise this exploration, the learning phase of the agent could start on the monthly model before being refined on the one with an hourly time-resolution, the so-called transfer learning \cite{mann2013directed}.  Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local VRES and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}). This section aims at comparing results from the learning process on the monthly model with the one on the hourly model. The monthly learning has been carried out by applying the same rules (see Section \ref{sec:RL:act_states_rew}). The purpose is to assess the capacity of the monthly model to be a good proxy of the hourly myopic transitions.

Considering the actions taken by the agent and their binding effect on the environment with a monthly resolution,  we see similar trends as in the hourly model (see Figure \ref{fig:app:Binding_constr_joss}). Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation. Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon-neutrality becomes more crucial at the end of the transition.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Binding_learning_TD_joss.pdf}
\includegraphics[width=0.49\textwidth]{Binding_learning_MO_joss.pdf}
\caption{Binding versus non-binding constraints on the hourly and monthly models. When the agent learns on the monthly model, the actions are binding in the similar timing and similar ranges as when learning on the hourly model. }
\label{fig:app:Binding_constr_joss}
\end{figure}

@@

In conclusion, with the advantage of the smaller computational burden, the monthly model myopic model is found to be a good proxy with its equivalent based on hourly time-resolution.

>>>>

The \acrfull{RL}-based exploration of myopic transition pathways can require thousands of runs to converge. To reduce the computational burden and maximise this exploration, the learning phase of the agent could start on the monthly model before being refined on the one with an hourly time resolution, the so-called transfer learning \cite{mann2013directed}.  Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local VRES and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}). This section aims at comparing results from the learning process on the monthly model with the one on the hourly model. The monthly learning has been carried out by applying the same rules (see Section \ref{sec:RL:act_states_rew}). The purpose is to assess the capacity of the monthly model to be a good proxy of the hourly myopic transitions.

Considering the actions taken by the agent and their binding effect on the environment with a monthly resolution,  we see similar trends as in the hourly model (see Figure \ref{fig:app:Binding_constr_joss}). Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation. Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon neutrality becomes more crucial at the end of the transition.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Binding_learning_TD_joss.pdf}
\includegraphics[width=0.49\textwidth]{Binding_learning_MO_joss.pdf}
\caption{Binding versus non-binding constraints on the hourly and monthly models: distribution of occurrences. When the agent learns on the monthly model, the actions are binding in the similar timing and similar ranges as when learning on the hourly model. }
\label{fig:app:Binding_constr_joss}
\end{figure}

@@

In conclusion, with the advantage of the smaller computational burden, the monthly model myopic model is found to be a good proxy with its equivalent based on hourly time resolution.

### 2024-07-19 appendices/app_case_study.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

\section{Belgian energy transition pathway towards carbon-neutrality in 2050} 
\label{app:bel_PF_TD}
This section presents the results of the deterministic (\ie all parameters at their nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon-neutrality in 2050.  After performing a technical investigation of the pathway by checking the greenhouse gas breakdown by energy sectors, the primary energy mix is analysed. To illustrate the sector coupling, a focus is made on the electrification of the sectors. Then, the cost implications in terms of investments and operations are discussed.

@@

 \begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{Gwp_per_sector.pdf}
\caption{Energy sectors have different rates to reduce \gls{GHG} emissions over the transition. The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon-neutrality in 2050.}
\label{fig:pestd_ghg}
\end{figure}

@@

\section{Assessment of different emissions-trajectories}
\label{app:CO2_trajectories}
This section compare results of the optimisation subject to different emissions-trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon-neutrality in 2050. The first comparison is done with the trajectory subject to respect the \ce{CO2}-budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}). Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, removing the constraint on the emissions-trajectory.

\subsection{\ce{CO2}-budget versus linear decrease of emissions}
\label{app:CO2_budget}

\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2}-budget) and a case where the \ce{CO2}-trajectory is constrained instead. Interestingly, these two transition pathways end up in a similar carbon-neutral whole-energy system in 2050. The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat. In the former, this is linked to the earlier phase-out of oil products through naphtha-cracking. In the latter, it is the use of coal in industrial coal boilers that is phased-out. Overall, ending up to the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{GWP_per_sector_lin.pdf}
\includegraphics[width=\textwidth]{GWP_per_sector_REF.pdf}
\caption{Respecting the \ce{CO2}-budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
\label{fig:app_CO2_REF_lin}
\end{figure}

@@

\begin{table}[htbp]
\caption{Comparison between the two different time-resolutions: Hourly (PF) and Monthly (MO). Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}
\label{tab:TD_MO}
\begin{minipage}{\textwidth}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{l l r r c}
\toprule
& & {\thead{\textbf{PF}}} & {\thead{\textbf{MO}}} & Units \\
\midrule
Computational time\footnote{\label{foot:comp_time_mo}These computational times were reached on a 2.4GHz 4-core machine.} & & 830 & \textbf{2} & s \\
\midrule
\multirow{4}{*}{Costs in 2050}
& Total transition\footnote{\label{foot:transition_cost_mo}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & -2\% & b€\\
& Cumulative opex & 528 & -3\%  & b€\\
& Cumulative capex & 636 & -2\% & b€\\
& Salvage value & 160 & -5\% & b€\\
\midrule
\multirow{5}{*}{Primary energy mix in 2050} & Total & 368.0 & +2\% & TWh/y\\
 & e-hydrogen & 15.6 & +2\% & TWh/y\\
 & e-methane & 41.0 & \textbf{+51\%} & TWh/y\\
 & e-methanol & 54.8 & $\simeq$ & TWh/y\\
 & e-ammonia & 40.7 & \textbf{+55\%} & TWh/y\\

>>>>

\section{Belgian energy transition pathway towards carbon neutrality in 2050} 
\label{app:bel_PF_TD}
This section presents the results of the deterministic (\ie all parameters at their nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon neutrality in 2050.  After performing a technical investigation of the pathway by checking the greenhouse gas breakdown by energy sectors, the primary energy mix is analysed. To illustrate the sector coupling, a focus is made on the electrification of the sectors. Then, the cost implications in terms of investments and operations are discussed.

@@

 \begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{Gwp_per_sector.pdf}
\caption{Energy sectors have different rates to reduce \gls{GHG} emissions over the transition. The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon neutrality in 2050.}
\label{fig:pestd_ghg}
\end{figure}

@@

\section{Assessment of different emissions-trajectories}
\label{app:CO2_trajectories}
This section compare results of the optimisation subject to different emissions-trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon neutrality in 2050. The first comparison is done with the trajectory subject to respect the \ce{CO2} budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}). Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, removing the constraint on the emissions-trajectory.

\subsection{\ce{CO2} budget versus linear decrease of emissions}
\label{app:CO2_budget}

\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2} budget) and a case where the \ce{CO2}-trajectory is constrained instead. Interestingly, these two transition pathways end up in a similar carbon neutral whole-energy system in 2050. The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat. In the former, this is linked to the earlier phase-out of oil products through naphtha-cracking. In the latter, it is the use of coal in industrial coal boilers that is phased-out. Overall, ending up to the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{GWP_per_sector_lin.pdf}
\includegraphics[width=\textwidth]{GWP_per_sector_REF.pdf}
\caption{Respecting the \ce{CO2} budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
\label{fig:app_CO2_REF_lin}
\end{figure}

@@

\begin{table}[htbp]
\caption{Comparison between the two different time resolutions: Hourly (PF) and Monthly (MO). Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}
\label{tab:TD_MO}
\begin{minipage}{\textwidth}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{l l r r c}
\toprule
& & {\thead{\textbf{PF}}} & {\thead{\textbf{MO}}} & Units \\
\midrule
Computational time\footnote{\label{foot:comp_time_mo}These computational times were reached on a 2.4GHz 4-core machine.} & & 830 & \textbf{2} & s \\
\midrule
\multirow{4}{*}{Costs in 2050}
& Total transition\footnote{\label{foot:transition_cost_mo}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & -2\% & b€\\
& Cumulative opex & 528 & -3\%  & b€\\
& Cumulative capex & 636 & -2\% & b€\\
& Salvage value & 160 & -5\% & b€\\
\midrule
\multirow{5}{*}{Primary energy mix in 2050} & Total & 368.0 & +2\% & TWh/y\\
 & e-hydrogen & 15.6 & +2\% & TWh/y\\
 & e-methane & 41.0 & \textbf{+51\%} & TWh/y\\
 & e-methanol & 54.8 & $\simeq$ & TWh/y\\
 & e-ammonia & 40.7 & \textbf{+55\%} & TWh/y\\
